---
title: "Streamflow Comparative Analysis"
author: "Emma Pingel"
date: "2024-03-21"
output: html_document
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

*Completed with project partner Lexi Bell for CSU's ESS 330:
Quantitative Reasoning for Ecosystem Science.*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Streamflow & Urbanization

As infrastructure increases and urbanization becomes more frequent,
natural landscapes are disrupted causing a series of negative domino
effects on surrounding ecosystems. This trend is especially relevant in
freshwater sources due to their sensitivity to change and pre-stressed
conditions that draw a significant amount of uncertainty in future
planning. To understand the extent of these impacts in Colorado, my
partner and I examined how streamflow might respond to nearby
infrastructure development using USGS data from two sites along the
Cache la Poudre River. Our null hypothesis assumed no difference in
streamflow between the two sites while our alternative hypothesis
suggests a significant difference in streamflow as the river intersects
with the built environment.

------------------------------------------------------------------------

#### Load in Data from USGS

Two sites were selected for this comparative analysis: Livermore and
Timnath. The prior is located northwest of Fort Collins and is
referenced as the site with no nearby development while Timnath is
downstream of the city. Using the `dataRetrieval` package, we can
download historical water quality data from USGS.

```{r}
# install.packages("dataRetrieval") if needed
library(dataRetrieval)

site_number_P1 <- "06751490" 
parameterCd_P1 <- "00060" 
Livermore <- readNWISuv(site_number_P1, parameterCd_P1, "2023-01-01", "2023-12-31")

site_number_P2 <- "06752280" 
parameterCd_P2 <- "00060" 
Timnath <- readNWISuv(site_number_P2, parameterCd_P2, "2023-01-01", "2023-12-31")
```

#### Tidy Data for Calculations

Since USGS water quality measurements are taken every 15 to 60 minutes,
it is crucial that they are summarized for the sake of comparison and
calculation. Focusing on the 2023 calendar year, we selected discharge
(cubic feet per second) as the is the main variable to illustrate
natural periods of higher/lower flow rates. Thus, the discharge rate at
each site will be averaged on monthly intervals.

```{r}
library(tidyverse)

# extract month from each row within dateTime column
Month1 = months(Livermore$dateTime)

# create new data frame with month and discharge only
discharge_data1 = data.frame(Month = Month1, Discharge = Livermore$X_00060_00000)

# remove NAs for higher quality data [not necessary but just in case]
discharge_data1 <- na.omit(discharge_data1)

# calculate monthly averages
discharge_avg1 <- group_by(discharge_data1, Month) %>%
  summarize(mon_discharge = mean(Discharge)) %>%
  select(Site = Month, mon_discharge)

# rename 'Month' to differentiate for dataframe binding
discharge_avg1$Site[c(1:12)] <- "Liv"


# repeat the same steps for the second site
Month2 = months(Timnath$dateTime)

# create new data frame with month and discharge only
discharge_data2 = data.frame(Month = Month2, Discharge = Timnath$X_00060_00000)

# remove NAs for higher quality data
discharge_data2 <- na.omit(discharge_data2)

# calculate monthly averages
discharge_avg2 <- group_by(discharge_data2, Month) %>%
  summarize(mon_discharge = mean(Discharge)) %>%
  select(Site = Month, mon_discharge)

# rename 'Month' to differentiate for dataframe binding
discharge_avg2$Site[c(1:12)] <- "Tim"
```

#### Combine Data Frames

For easier calculations, the Livermore and Timnath data frames were
binded together.

```{r}
library(dplyr)

# merge data frames by row
row_avg_both <- bind_rows(discharge_avg1, discharge_avg2)
```

#### Run T-Test to Compare Sites

Following the steps for a typical t-test, the normality and equal
variance assumptions are first tested and then corrected. A paired
t-test is the default model with confirmation from Welch's t-test.

```{r}
# visually test for normality
hist(discharge_avg1$mon_discharge)
hist(discharge_avg2$mon_discharge)
```

Both Livermore and Timnath's discharge distributions are skewed to the
right; thus, a log transformation is necessary to force data into
normality.

```{r}
# visually test for equal variance
row_avg_both %>%
  ggplot(aes(x = Site, y = mon_discharge)) +
  geom_boxplot()
```

The central tendencies are similar but not great indicators since
Timnath has a much larger range of values. Instead, we will confirm the
equal variance assumption with a mathematical test.

```{r}
# 'pull' variables
Liv_avg <- row_avg_both %>%
  filter(Site == "Liv") %>%
  pull(mon_discharge)

Tim_avg <- row_avg_both %>%
  filter(Site == "Tim") %>%
  pull(mon_discharge)

# test EV asusmption
var.test(Liv_avg, Tim_avg)
```

Since the p-value (8.644e-05) is less than 0.05, we can assume that the
variables do not have relatively equal intervals and a log
transformation is definitely needed.

```{r}
# log transform necessary due to unequal variance & non normality
var.test(log(Liv_avg), log(Tim_avg))
```

Because the p-value (0.253) is not greater than 0.05, we know that the
variance and normality of the data have been corrected so we can proceed
with the t-test.

```{r}
# perform paired t-test
t.test(log(row_avg_both$mon_discharge) ~ row_avg_both$Site, var.equal = T)

# perform unpaired t-test (welch) to confirm answer
t.test(row_avg_both$mon_discharge ~ row_avg_both$Site, var.equal = F)
```

The p-value (0.474) in the paired t-test is greater than 0.05 so we fail
to reject the null hypothesis and conclude that there is not a
significant difference in discharge rates at the two sites representing
urbanization's impact. Using an unpaired t-test, the p-value (0.262) is
also larger than 0.05 so we again fail to reject the null hypothesis and
conclude that the two sites do not have significant variation in
discharge.

#### Visualization

For the purpose of our final report, my partner and I spent some time
creating a scatterplot to visualize our findings. This required the
individual sites' data frames to be rejoined by column and labeled by
month.

```{r}
#change labels to create scatterplot for monthly averages
Month1 = months(Livermore$dateTime)
discharge_data1 = data.frame(Month = Month1, Discharge = Livermore$X_00060_00000)
discharge_data1 <- na.omit(discharge_data1)
discharge_avg1 <- group_by(discharge_data1, Month) %>%
  summarize(mon_discharge = mean(Discharge))

Month2 = months(Timnath$dateTime)
discharge_data2 = data.frame(Month = Month2, Discharge = Timnath$X_00060_00000)
discharge_data2 <- na.omit(discharge_data2)
discharge_avg2 <- group_by(discharge_data2, Month) %>%
  summarize(mon_discharge = mean(Discharge))

#combine dataframes by column
col_avg_both <- merge.data.frame(discharge_avg1, discharge_avg2, by = "Month")

#recombine dataframes by row including month
row2_avg_both <- bind_rows(discharge_avg1, discharge_avg2)
ggplot_data <- merge.data.frame(row_avg_both, row2_avg_both, by = "mon_discharge")


#use ggplot to generate figure
ggplot_data %>%
  ggplot(aes(x = Month, y = mon_discharge)) +
  geom_point(aes(color = Site)) +
  ggtitle("Discharge Rates at Livermore & Timnath") +
  xlab("Month") +
  ylab("Avg Discharge (cubic ft/sec)") +
  theme_minimal()
```

To conclude, our initial prediction assumed that there would be a
relationship between urbanization and streamflow dynamics; however, we
ended up failing to reject our null hypothesis. Thus, we did not find a
significant difference between streamflow in non-developed vs developed
settings and do not believe there is an obvious relationship between the
two variables. Although this outcome didn’t correlate with our
prediction, it did match the findings of similar case studies
experimenting with urbanization and aquatic ecosystems.

While we averaged discharge rates by month to minimize confounding bias,
we also want to recognize the challenge that our temporal scale
presented in looking at one year of data. For a more inclusive analysis,
we would likely want to repeat the t-tests with other years or include
them in our initial test using an analysis of variance test (ANOVA)
instead.
